Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated

### 2015-04-24 manuscript.Rmd
equations, typesetting, some code

```{r echo=FALSE, message=FALSE}
library(knitr)
load("cache/lai_workspace.RData")
source("load.R")
```

@@

The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated (Crous et al., 2015). Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees ha-1. The vegetation consists of an open woodland (basal area before treatment, BA = 27.6 ± 2.7 m$^2$ ha‑1, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by Microlaena stipoides (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum (Hendrey et al., 1999) through which a computer-controlled system modulates release of diluted CO$_2$ into the vegetation according to a proportional-integral-differential type of algorithm (Lewin et al., 2009). The vegetation within three randomly selected rings was exposed to a CO$_2$ mole fraction (C$_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO$_2$ treatment rings in infrastructure and instrumentation but received ambient C$_a$, and are used as control plots. The target diurnal C$_a$ for the rings exposed to e*C*$_a$ was 150 μmol CO$_2$ mol‑1 above ambient C$_a$ in full operation mode. However, C$_a$ was increased gradually for six months at a rate of ca. 30 μmol CO$_2$ mol-1 month-1 until full operation mode was reached (Figure 1a).

Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP ® sensor (HMP 155 Vaisala, Vantaa, Finland). Total precipitation was monitored with three automated tipping buckets (Tipping Bucket Rain gauge TB4, Hydrological Services Pty Ltd, Liverpool, NSW, Australia) located at the top of the central tower at 23.5 m height, in three of the rings. Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers. Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50 depth depending on the ring).

@@

"*L*" _p=-loga(τ_p)/k									(1)

@@

τ_d=Q ̅_below/Q_above								(2)

where τ_d is diffuse canopy transmittance, Qabove is above canopy PAR (µmol m$^{-2}$ s-1), and Q ̅_below is the average PAR across the three sub-canopy PAR sensors in each FACE ring. We then calculated daily averages of τ_d, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was  > 0.03 (in units of τ_d).

@@

τ_d=2∫_0^(π/2)▒〖e^(-LAI∙G(θ)/cosa(θ))  sina(θ)  cosa(θ)dθ〗						(3)

@@

To overcome violations in the assumptions underlying Eq. 3, we calibrated *L* against litter production. We estimated *L* from τ_d (*L*τ) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig. 1). We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus τ_d increased steadily, see Fig. 1). During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period). A total of n=11 days with sufficiently diffuse conditions to estimate τ_d  were available during this period. In theory, cumulative *L* leaf litter production (PL, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance. We calibrated equation (4),

c〖(LAI〗_τ-〖LAI〗_(t=0))=∫_(t=0)^t▒〖P_L dt〗								(4)

where c is a calibration parameter, Lt=0 is *L*τ at the start of the drought period. We used linear regression without intercept to estimate c in Eq. 4 as 0.68 (SE = 0.035). It should be stressed that even after calibration with litter production, *L*τ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq. 4.

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and τd timeseries (Wood, 2006), using the mgcv package in R (R Core Team, 2014). For all time-series, we used a base of 18 when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson (2014). For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of C$_a$ treatment on *L*τ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C$_a$ treatment and date as a factor.

To assess effects of e*C*$_a$ on *L*τ, we accounted for pre-treatment differences in canopy structure, as follows. The average *L*τ over the entire period was well correlated with basal area (BA) measured prior to the onset of the e*C*$_a$ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to C$_a$ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual *L*τ from the fitted relationship with BA shown in Fig. 3, to yield a ‘*L* anomaly’. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

Canopy transmittance (τ_d) varied between 0.10 and 0.36, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig. 1).  However, temporal trends were very similar between the rings. Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only.

An outbreak of psyllids (Cardiaspina sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept. 2014 (for description of a similar outbreak in Eucalyptus moluccana, see (Hall et al., 2015)). This rapid decrease in *L* was evident in a quick increase in τ_d (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower τ_d than before the insect attack.

We compared calibrated *L*τ to an independent estimate of *L* from canopy photos (Lp), which were collected nine times between Oct. 2012 and Oct. 2013. Without further calibration, the two estimates were reasonably well correlated, though the canopy photos gave slightly higher estimates of *L* (Fig. 2, R2=0.71, RMSE=0.12). This is encouraging because the photos measure *L* near the zenith (+/- 15°), while *L*τ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

We found no differences in the *L* anomaly between the C$_a$ treatments (Fig. 4) (p > 0.05), since the 95% confidence intervals of the generalized additive model overlapped. A linear mixed-effects model with C$_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C$_a$ on the *L* anomaly (P > 0.2). What difference between C$_a$ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was 0.047 m$^2$ m$^{-2}$ for a*C*$_a$, and 0.042 m$^2$ m$^{-2}$ for the e*C*$_a$ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of 0.089 m$^2$ m$^{-2}$ (the sum of the half-width CI’s), or about a change of 6% over the mean *L* of the EucFACE (1.39 m$^2$ m$^{-2}$).

The dynamics of *L*τ were compared with litter fall, by estimating the absolute change in *L*τ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to *L*τ for each ring separately (similar to shown for τd in Fig. 1), and calculating the change in *L*τ for each period (Δ*L*τ). Litter production correlated significantly with Δ*L*τ when the latter was separated into periods where Δ*L*τ > 0 (net leaf growth) and Δ*L*τ <  0 (net leaf loss) (Fig. 5). During periods of net leaf loss, we found a significant correlation (with a slope of XX ± se, P < 0.001?, df or n?) close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing. During periods of net leaf growth, we found a significant correlation (slope, P, df…) between Δ*L*τ and litter production (Fig. 5).

>>>>

```{r echo=FALSE, message=FALSE}
library(knitr)
load("cache/lai_workspace.RData")
source("load.R")
library(broom)
options(scipen = 1, digits = 2)

# Some pre-calculations
caliblm <- calibrateToDrought(facegap_cloudy_byring)$lmfit

# Some definitions of constants
.gambase <- 18
```

@@

The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated (Crous et al., 2015). Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees ha-1. The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m$^2$ ha$^{-1}$, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by Microlaena stipoides (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum (Hendrey et al., 1999) through which a computer-controlled system modulates release of diluted CO$_2$ into the vegetation according to a proportional-integral-differential type of algorithm (Lewin et al., 2009). The vegetation within three randomly selected rings was exposed to a CO$_2$ mole fraction (C$_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO$_2$ treatment rings in infrastructure and instrumentation but received ambient C$_a$, and are used as control plots. The target diurnal C$_a$ for the rings exposed to e*C*$_a$ was 150 μmol CO$_2$ mol$^{-1}$ above ambient C$_a$ in full operation mode. However, C$_a$ was increased gradually for six months at a rate of ca. 30 μmol CO$_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Figure 1a).

Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP sensor (HMP 155 Vaisala, Vantaa, Finland). Total precipitation was monitored with three automated tipping buckets (Tipping Bucket Rain gauge TB4, Hydrological Services Pty Ltd, Liverpool, NSW, Australia) located at the top of the central tower at 23.5 m height, in three of the rings. Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers. Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50cm depth depending on the ring).

@@

(1) $$L_p = -log(\tau_p)/k$$

@@

(2) $$\tau_d = \bar{Q}_{below} / Q_{above}$$

where $\tau_d$ is diffuse canopy transmittance, Qabove is above canopy PAR (µmol m$^{-2}s^{-1}$), and Q$_{below}$ is the average PAR across the three sub-canopy PAR sensors in each FACE ring. We then calculated daily averages of $\tau_d$, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was  > 0.03 (in units of $\tau_d$).

@@

(3) $$\tau_d = 2\int_0^{\pi/2} e^{-LG(\theta)/cos(\theta)}sin(\theta)cos(\theta)d\theta$$

@@

To overcome violations in the assumptions underlying Eq. 3, we calibrated *L* against litter production. We estimated *L* from $\tau_d$ ($L_\tau$) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig. 1). We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period). A total of n=11 days with sufficiently diffuse conditions to estimate τ_d  were available during this period. In theory, cumulative *L* leaf litter production (PL, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance. We calibrated equation (4),

(4) $$c(L_{\tau,t} - L_{\tau,t=0}) = \int_0^t P_Ldt$$

where *c* is a calibration parameter, Lt=0 is $L_\tau$ at the start of the drought period. We used linear regression without intercept to estimate c in Eq. 4 as `r calib` (SE = `r tidy(caliblm)$std.error`). It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq. 4.

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and τd timeseries (Wood, 2006), using the mgcv package in R (R Core Team, 2014). For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson (2014). For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of C$_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C$_a$ treatment and date as a factor.

To assess effects of e*C*$_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows. The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the e*C*$_a$ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to C$_a$ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig. 3, to yield a ‘*L* anomaly’. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

Canopy transmittance ($\tau_d$) varied between 0.10 and 0.36, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig. 1).  However, temporal trends were very similar between the rings. Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only.

An outbreak of psyllids (Cardiaspina sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept. 2014 (for description of a similar outbreak in Eucalyptus moluccana, see (Hall et al., 2015)). This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower τ_d than before the insect attack.

We compared calibrated $L_\tau$ to an independent estimate of *L* from canopy photos ($L_p$), which were collected nine times between Oct. 2012 and Oct. 2013. Without further calibration, the two estimates were reasonably well correlated, though the canopy photos gave slightly higher estimates of *L* (Fig. 2, R2=0.71, RMSE=0.12). This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

We found no differences in the *L* anomaly between the C$_a$ treatments (Fig. 4) (p > 0.05), since the 95% confidence intervals of the generalized additive model overlapped. A linear mixed-effects model with C$_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C$_a$ on the *L* anomaly (P > 0.2). What difference between C$_a$ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was 0.047 m$^2$ m$^{-2}$ for a*C*$_a$, and 0.042 m$^2$ m$^{-2}$ for the $eC_a$ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of 0.089 m$^2$ m$^{-2}$ (the sum of the half-width CI’s), or about a change of 6% over the mean *L* of the EucFACE (1.39 m$^2$ m$^{-2}$).

The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to shown for $\tau_d$ in Fig. 1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$). Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 5). During periods of net leaf loss, we found a significant correlation (with a slope of XX ± se, P < 0.001?, df or n?) close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing. During periods of net leaf growth, we found a significant correlation (slope, P, df…) between $\Delta L_\tau$ and litter production (Fig. 5).
